Two Problems for Sophistication

### 2015-01-21 Facticity/Facticity/paper.tex
...

In the previous section, we saw the first glimpse of the delicate balance between the two code components in sophistication. If the description of the model $m$ is less efficient than $C^\K(m)$ by anything more than a constant, the sophistication becomes bounded. We will investigate this balance, starting with the sophistication $S^{\K,\psi,c}$. In this setting, we know that, on the one hand, there always exist candidate models---those models $i$ with $i \in \K$ and $C^{i}(x) \leq C^\K(x) + c$---for which al information except a constant amount is stored in the data, and on the other hand, there are also candidate representations for which all information is stored in the model. This tells us that $S^\K$ is \emph{balanced}: there is no inefficiency in the representation forcing information in or out of the model.
The flip side of having a balanced sophistication is that it becomes easy to show invariance. We will show that there are numberings for which the universal models always determine the $S^\K$ and numberings for which the singletons always determine $S^\K$. The following lemma is based on the idea that, given a subset $\M'$ of the model class, we can choose the numbering so that representations use a model in $\M'$ perform an arbitrary constant better than all other models, so that a model in $\M'$ always ends up determining the sophistication.
\begin{restatable}{lemma}{coolone}
\label{lemma:thecoolone}
  Let $\M$ be any model class, let $\X$ be any set of binary strings and let $D:\B\to\N$ be a computable decoding function with a prefix-free domain that maps function descriptions to their indices in $\phi$, i.e. if $m=\phi_{D(p)}$ then $p$ is a $D$-description of $m$. Let $\M'=\{\phi_i:i\in\tn{range}(D)\}$. Further assume there is a constant $c$ such that:\\
\-\hspace{1cm}(1) $\forall_{m\in\M'}:\min\{|p|:\phi_{D(p)}=m\}\le C^{\K,\phi}(m)+c$\\
\-\hspace{1cm}(2) $\forall_{x\in\X}:C^{\M',\phi}(x)-C^{\M,\phi}(x)\le c$.\\
Then there is an enumeration $\psi$ of the partial recursive functions such that $S^{\M,\psi}(x) = |\br{0}|+S^{\M',\phi}(x)$ for all $x\in\X$.
\end{restatable}
This means that for model classes with a universal model, there is a numbering such that sophistication is bounded:
\begin{restatable}[underfitting]{lemma}{underfitting}
Let $\M$ be a model class containing a universal model $\phi_u$, with the property that $\exists c \forall i \in \M : C^{\{u\}}(x) \leq C^{\{i\}}(x) + c \p$. Then, there exist numberings such that $\s^\M$ is bounded.
\end{restatable}
\begin{restatable}[Overfitting]{lemma}{overfitting}
Let $\M \subseteq \K$ be a model class where for every $x\in\X$ there is a singleton model $i\in\M$ with $\phi_i(\epsilon)=x$. Then there is a numbering $\psi$, and a constant $c$, such that for all $x\in\X$ we have $C^\M(x)-S^{\M,\psi}(x)\leq c$.\label{lemma:overfitting}
\end{restatable}
\noindent Thus, in this balanced sophistication, there is no invariance: all information can be seen as structure, or as noise, depending on the numbering. To avoid these issues, existing proposals upset the balance to exclude or penalize at least the singleton models and the universal models.

@@

Thus, there are no longer any numberings for which sophistication is bounded, so the underfitting problem is at least partly solved. However, as we show below, the resulting sophistication is not an intuitive measure of the amount of structure in the data. Moreover, we push the balance back in the direction of overfitting, which is only exacerbated by this change. Only one model class eliminates both the singletons and the universal model: $\T$. The only proposal that uses an efficient model representation \emph{and} excludes the universal models \emph{and} excludes the singletons is Vereshchagin and Vit\'anyi's \cite{vitanyi2004meaningful}. While this invalidates straightforward proofs of boundedness, there is no suggestion that $S^{\T, \phi, c}(x)$ is actually invariant.

>>>>

In the previous section, we saw the first glimpse of the delicate balance between the two code components in sophistication. If the description of the model $m$ is less efficient than $C^\K(m)$ by anything more than a constant, the sophistication becomes bounded. We will investigate this balance, starting with the sophistication $S^{\K,\psi,c}$. In this setting, we know that, on the one hand, there always exist candidate models---those models $i$ with $i \in \K$ and $C^{i}(x) \leq C^\K(x) + c$---for which al information except a constant amount is stored in the data, and on the other hand, there are also candidate representations for which all information is stored in the model. This tells us that $S^\K$ is \emph{balanced}: there is no inefficiency in the representation forcing information in or out of the model.
The flip side of having a balanced sophistication is that it becomes easy to show invariance. We will show that there are numberings for which the universal models always determine the $S^\K$ and numberings for which the singletons always determine $S^\K$. The following lemma is based on the idea that, given a subset $\M'$ of the model class, we can choose the numbering so that representations use a model in $\M'$ perform an arbitrary constant better than all other models, so that a model in $\M'$ always ends up determining the sophistication.
\begin{restatable}{lemma}{coolone}
\label{lemma:thecoolone}
  Let $\M$ be any model class, let $\X$ be any set of binary strings and let $D:\B\to\N$ be a computable decoding function with a prefix-free domain that maps function descriptions to their indices in $\phi$, i.e. if $m=\phi_{D(p)}$ then $p$ is a $D$-description of $m$. Let $\M'=\{\phi_i:i\in\tn{range}(D)\}$. Further assume there is a constant $c$ such that:\\
\-\hspace{1cm}(1) $\forall_{m\in\M'}:\min\{|p|:\phi_{D(p)}=m\}\le C^{\K,\phi}(m)+c$\\
\-\hspace{1cm}(2) $\forall_{x\in\X}:C^{\M',\phi}(x)-C^{\M,\phi}(x)\le c$.\\
Then there is an enumeration $\psi$ of the partial recursive functions such that $S^{\M,\psi}(x) = |\br{0}|+S^{\M',\phi}(x)$ for all $x\in\X$.
\end{restatable}
This means that for model classes with a universal model, we can choose our numbering to make $S^\K$ bounded, or equal to the complexity.
\begin{restatable}[underfitting]{lemma}{underfitting}
Let $\M$ be a model class containing a universal model $\phi_u$, with the property that $\exists c \forall i \in \M : C^{\{u\}}(x) \leq C^{\{i\}}(x) + c \p$. Then, there exist numberings such that $\s^\M$ is bounded.
\end{restatable}
\begin{restatable}[Overfitting]{lemma}{overfitting}
Let $\M \subseteq \K$ be a model class where for every $x\in\X$ there is a singleton model $i\in\M$ with $\phi_i(\epsilon)=x$. Then there is a numbering $\psi$, and a constant $c$, such that for all $x\in\X$ we have $C^\M(x)-S^{\M,\psi}(x)\leq c$.\label{lemma:overfitting}
\end{restatable}
\noindent Thus, in this balanced sophistication, there is no invariance: all information can be seen as structure, or as noise, depending on the numbering. To avoid these issues, existing proposals upset the balance to exclude or penalize at least the singleton models and the universal models.

@@

Thus, there are no longer any numberings for which sophistication is bounded. However, as we show below, the resulting sophistication is not an intuitive measure of the amount of structure in the data. Moreover, the balance is pushed back in the direction of overfitting, which is only exacerbated by this change. Only one model class eliminates both the singletons and the universal model: $\T$. The only proposal that uses an efficient model representation \emph{and} excludes the universal models \emph{and} excludes the singletons is Vereshchagin and Vit\'anyi's \cite{vitanyi2004meaningful}. While this invalidates straightforward proofs of boundedness, there is no suggestion that $S^{\T, \phi, c}(x)$ is actually invariant.
